---
title: "analysis"
format: html
editor: source
---

## Setup

```{r}
#| message: false
library(tidyverse)
library(haven)
library(sjmisc)
```

## Loading in data

```{r}
# insert your own file names for the datasets here
round1 <- read_dta(file.path("data", "ggp", "round1.dta"))
round2 <- read_dta(file.path("data", "ggp", "round2.dta"))
pilot <- read_dta(file.path("data", "ggp", "pilot.dta"))
```

## Tidy round 1

```{r}
round1tidy <- round1 |>
  select(
    aparstat, amarstat, a1107_a, a332, a1002, a301m, a301y, a302bm, a302by,
    a148, a308, asex, a384, abyear, a373m, a373y, aactstat, arid, acountry,
    amonth, ayear, aweight
  ) |>
  mutate(
    parstat = case_when(
      aparstat == 1 & (amarstat == 1 | amarstat == 3 | amarstat == 4) ~ 1, # cohabiting
      aparstat == 1 & amarstat == 2 ~ 2,                                    # married
      aparstat == 2 ~ 3,                                                    # LAT
      aparstat == 3 ~ 4,                                                    # single
      TRUE ~ NA
    )
  ) |>
  # harmonize variable names with later waves
  rename(
    att03a = a1107_a,
    dem28c = a332,
    inc03  = a1002,
    edu    = a148,
    edup   = a308,
    sex    = asex,
    sexp   = a384,
    birthy = abyear,
    birthpm = a373m,
    birthpy = a373y,
    activ  = aactstat,
    respid = arid,
    weight = aweight
  )
```

## Tidy round 2 and pilot

Round 2 and the pilot share the same variable structure and transformations. The only
difference is the education variable codes (1–8 in round 2, 2801–2808 in the pilot).

```{r}
tidy_r2_format <- function(df, edu_codes) {
  df |>
    mutate(
      parstat = case_when(
        dem21 == 1 & dem28a == 2 & dem30a == 1 ~ 1, # cohabiting
        dem21 == 1 & dem28a == 1 ~ 2,               # married
        dem21 == 1 & dem28a == 2 & dem30a == 2 ~ 3, # LAT
        dem21 == 2 ~ 4,                              # single
        TRUE ~ NA
      ),
      dem07 = case_match(dem07,
        edu_codes[1:2] ~ 2, edu_codes[3] ~ 3, edu_codes[4] ~ 4,
        edu_codes[5:6] ~ 5, edu_codes[7:8] ~ 6, .default = NA
      ),
      dem25 = case_match(dem25,
        edu_codes[1:2] ~ 2, edu_codes[3] ~ 3, edu_codes[4] ~ 4,
        edu_codes[5:6] ~ 5, edu_codes[7:8] ~ 6, .default = NA
      ),
      dem06 = case_match(dem06,
        c(2, 3) ~ 1, 4 ~ 2, 5 ~ 3, 1 ~ 4, 6 ~ 5,
        c(9, 10) ~ 6, 11 ~ 7, 7 ~ 9, 12 ~ 10, .default = NA
      )
    ) |>
    rename(
      a301m   = dem30bm, a301y   = dem30by,
      a302bm  = dem28bm, a302by  = dem28by,
      edu     = dem07,   edup    = dem25,
      sex     = dem01,   sexp    = dem23,
      birthm  = dem02m,  birthy  = dem02y,
      birthpm = dem22m,  birthpy = dem22y,
      activ   = dem06,
      acountry = country,
      amonth  = intdatem, ayear  = intdatey
    )
}

round2tidy <- round2 |>
  select(
    dem21, dem28a, dem30a, dem28c, att03a, inc03, dem30bm, dem30by, dem28bm, dem28by,
    dem33, dem07, dem25, dem01, dem23, dem02m, dem02y, dem22m, dem22y, dem06,
    respid, country, mode, intdatem, intdatey, weight
  ) |>
  tidy_r2_format(edu_codes = 1:8)

# pilot has no weight variable; weight is set to 1 in the append step
pilot2tidy <- pilot |>
  select(
    dem21, dem28a, dem30a, dem28c, att03a, inc03, dem30bm, dem30by, dem28bm, dem28by,
    dem33, dem07, dem25, dem01, dem23, dem02m, dem02y, dem22m, dem22y, dem06,
    respid, country, mode, intdatem, intdatey
  ) |>
  tidy_r2_format(edu_codes = 2801:2808)
```

## Appending

```{r}
#| message: false
#| warning: false
analysis <- add_rows(round1tidy, round2tidy, pilot2tidy, id = "survey_type") |>
  mutate(
    survey = if_else(survey_type == 1, "2005", "2021"),
    weight = if_else(survey_type == 3, 1, weight)
  )
```

## Creation of typology

Key variables in construction of typology:

- `att` — marriage is outdated
- `int3` — plans to marry within 3 years
- `econ` — household able to make ends meet

```{r}
analysis <- analysis |>
  mutate(
    # 2005 scale runs agree→disagree; reverse it so all waves run disagree→agree
    att03afix = if_else(survey_type != 1, 6 - att03a, att03a),

    att = case_when(
      att03afix == 1 | att03afix == 2 ~ 1, # agree
      att03afix == 4 | att03afix == 5 ~ 2, # disagree
      att03afix == 3 ~ 3,                  # indifferent
      TRUE ~ NA
    ),

    # 2021 adds an "unsure" option (value 3) that 2005 lacks;
    # shift 2005 values 3 and 4 up by 1 to align the scales
    dem28cfix = if_else(survey_type == 1 & dem28c > 2, dem28c + 1, dem28c),

    int3 = case_when(
      dem28cfix == 4 | dem28cfix == 5 ~ 1,                  # yes
      dem28cfix == 1 | dem28cfix == 2 | dem28cfix == 3 ~ 0, # no
      TRUE ~ NA
    ),

    econ = case_when(
      inc03 == 1 | inc03 == 2 ~ 1,                             # yes
      inc03 == 3 | inc03 == 4 | inc03 == 5 | inc03 == 6 ~ 0,  # no
      TRUE ~ NA
    ),

    cohab = if_else(parstat == 1, 1, 0),
    cores = if_else(parstat == 1 | parstat == 2, 1, 0),
    relat = if_else(parstat == 1 | parstat == 2 | parstat == 3, 1, 0),

    edu3 = case_when(
      edu == 2 ~ "low",
      edu %in% c(3, 4) ~ "mid",
      edu %in% c(5, 6, 7, 8) ~ "high",
      is.na(edu) ~ "unknown",
      TRUE ~ NA_character_
    ),

    age = ayear - birthy,

    age3 = case_when(
      age >= 18 & age <= 31 ~ "18-31",
      age >= 32 & age <= 45 ~ "32-45",
      age >= 46 & age <= 59 ~ "46-59",
      TRUE ~ NA_character_
    ),

    age4 = case_when(
      age >= 18 & age <= 34 ~ "18-34",
      age >= 35 & age <= 51 ~ "35-51",
      age >= 52 & age <= 69 ~ "52-69",
      age >= 70 ~ "70+",
      TRUE ~ NA_character_
    ),

    typ = case_when(
      cohab==1 & int3==1 & att==2              ~ "Prelude",
      cohab==1 & int3==0 & att==2 & econ==0   ~ "Trial",
      cohab==1 & int3==0 & att==2 & econ==1   ~ "Economic",
      cohab==1 & int3==1 & att %in% c(1, 3)   ~ "Conformist",
      cohab==1 & int3==0 & att==1              ~ "Refusal",
      cohab==1 & int3==0 & att==3              ~ "Irrelevant",
      TRUE ~ NA_character_
    ),

    big_type = case_when(
      typ %in% c("Prelude", "Conformist")    ~ "Forerunners - most likely getting married",
      typ %in% c("Trial", "Economic")        ~ "Forerunners - possibly getting married",
      typ %in% c("Refusal", "Irrelevant")    ~ "Alternatives - unlikely to get married",
      TRUE ~ NA_character_
    )
  )
```

## Analysis of typologies

```{r}
analysis_sample <- analysis |>
  # filter(survey_type != 3) |>  # optional: exclude pilot
  filter(!is.na(parstat), age < 70) |>
  filter(edu3 %in% c("low", "mid", "high"), !is.na(typ))

figure4 <- analysis_sample |>
  summarise(
    weight = sum(weight),
    .by = c("survey", "big_type", "typ")
  ) |>
  mutate(total_weight = sum(weight), .by = survey) |>
  mutate(prop = weight / total_weight) |>
  mutate(weight_big = sum(weight), .by = c("survey", "big_type")) |>
  mutate(prob_big = weight_big / total_weight) |>
  arrange(survey, big_type, typ)

figure5 <- analysis_sample |>
  summarise(
    weight = sum(weight),
    .by = c("survey", "age4", "typ")
  ) |>
  mutate(total_weight = sum(weight), .by = c("survey", "age4")) |>
  mutate(prop = weight / total_weight) |>
  arrange(survey, typ, age4)

figure6 <- analysis_sample |>
  summarise(
    weight = sum(weight),
    .by = c("survey", "edu3", "typ")
  ) |>
  mutate(total_weight = sum(weight), .by = c("survey", "edu3")) |>
  mutate(prop = weight / total_weight) |>
  arrange(survey, typ, edu3)
```
